---
title: "An introduction to SCM: Fit an additive covariance matrix model for multivariate normal data "
author: "Vincenzo Gioia and Matteo Fasiolo"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    number_sections: true
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{SCM}
  %\VignetteEncoding{UTF-8}
---

<style>
body {
text-align: justify}
</style>

```{r setup, include=FALSE}
library(knitr)
library(rgl)
opts_chunk$set(out.extra='style="display:block; margin: auto"', fig.align="center", tidy=FALSE)
knit_hooks$set(webgl = hook_webgl)
if (!requireNamespace("rmarkdown", quietly = TRUE) ||
!rmarkdown::pandoc_available("1.14")) {
warning(call. = FALSE, "These vignettes assume rmarkdown and pandoc
version 1.14. These were not found. Older versions will not work.")
knitr::knit_exit()
}
```

# Multivariate Gaussian additive models
Let  $\boldsymbol y_i\sim \mathcal{N}(\boldsymbol \mu_i, \boldsymbol \Sigma_i)$, $i=1, \ldots, n$, be independent response vectors. 

The covariates enter the model through the linear predictor vector $\boldsymbol \eta_i=(\eta_{i1}, \ldots, \eta_{iq})$, with $q= d+d(d+1)/2$, whose components allow specifying linear and smooth effects. 

The mean model is specified by $\mu_{ij} = \eta_{ij}$, $j=1, \ldots,d$, while the remaining $\eta_{ij}$, $j=d+1, \ldots,q$, specify the nonredundant, unconstrained elements of a covariance matrix parametrisation. The latter can be stated via the  modified Cholesky decomposition (MCD) by Pourahmadi (1999) or the matrix logarithmic tranformation by Chiu et al. (1996) 


The multivariate Gaussian additive models are fitted via the `gam_scm()` wrapper
```{r fit_mcd, echo=TRUE, eval=FALSE}
gam_scm(formula, family = mvn_scm(d = 2, param = NULL, nb = 1), optimizer = NULL, data = list(), aGam = list())
```


The inputs of the `gam_scm()` function are:

* `formula`: list of model formulas;
* `mvn_scm`: model family:
    - `d`: dimension of the outcome vector;
    - `param`: by default the MCD parametrisation (`"mcd"`), otherwise set (`"logm"`) for the logM one; 
    - `nb`: number of of blocks for chunking the evaluation of some quantities (default no chunks, while consider to increase such value if memory problem appears);
* `data`: a data frame or a list containing the outcomes and the covariates; 
* `optimizer`: optional argument specifying the approach to smoothing parameter optimisation; by default the generalised Fellner-Schall method (`"efs"`) by Wood and Fasiolo (2017), otherwise *only for the MCD parametrisation* the quasi Newton's optimisation  of the Laplace approximate marginal likelihood criterion (`"bfgs"`) according to Wood et al. (2016); 
* `aGam`: a list of auxiliary arguments that could be passed to the `gam()` function of the **mgcv** package (Wood, 2017). 


# Illustration
We introduce the most important functions of the package by means of the  GefCom2014 dataset (Hong et al., 2016). For illustrative purposes, the hourly loads (in MW) from 5 p.m. (`load_h17`) to 8 p.m (`load_h20`)  on a daily basis are considered as the outcome variables. The dataset, spanning the period from 2005/01/02 to 2011/11/30, includes the day of the year (`doy`), the day of the week (`dow`),  the exponentially smoothed temperature  (`temp95_h<k>`) and the loads of the previous day (`load24_h<k>`) at the $k$-th hour, $k=17,\ldots, 20$. 


```{r loadingData, echo=TRUE}
library(SCM)
data(GEF14_d4)
d <- 4
```



We start specifying the model formula for the mean vector: 

```{r meanmodelformula, echo=TRUE}
my_k = 15
my_bs = "cr"
mformula <- list(load_h17 | load_h18 | load_h19 | load_h20  ~ dow + s(doy, k = my_k, bs = my_bs),
                 load_h17 ~ load24_h17 + s(temp95_h17),
                 load_h18 ~ load24_h18 + s(temp95_h18),
                 load_h19 ~ load24_h19 + s(temp95_h19),
                 load_h20 ~ load24_h20 + s(temp95_h20))
```

The model formula is composed by two parts:

* "Common" specification:  by using the `|` operator, all the outcomes of the l.h.s. of `y ~ x` involve the same covariates  (first row);
* "Single-element" specification: the (rest of the) formula involving element-specific covariates can be stated separately (last 4 rows);


For simplicity, we consider the MCD-based additive covariance matrix model, which means modelling the elements of $\log \boldsymbol D^2_i$ and $\boldsymbol T_i$ resulting from $\boldsymbol \Sigma^{-1}_i={\boldsymbol T^\top_i}{ \boldsymbol D^{-2}_i} {\boldsymbol T_i}$.  For this application, it means to specify a model formula for  the non-trivial elements of    

$$ \log \boldsymbol D^2_i = \begin{pmatrix}\eta_{i5}  & 0 & 0  & 0 \\ 0& \eta_{i6} & 0 & 0 \\0  & 0 & \eta_{i7} & 0 \\0 & 0 & 0 & \eta_{i8}\end{pmatrix} \hspace{2cm} \boldsymbol T_i= \begin{pmatrix} 1  & 0 & 0 & 0 \\ \eta_{i9}& 1 & 0 & 0 \\ \eta_{i10}  &  \eta_{i11} & 1 & 0 \\\eta_{i12} & \eta_{i13} & \eta_{i14} &  1\end{pmatrix}$$

We facilitate the model formula specification (implicitly allowing a connection with the logM parametrisation) by introducing a matrix $\boldsymbol \Theta_i$ as in the form 

$$\boldsymbol \Theta_i= \begin{pmatrix} \eta_{i4}  & \eta_{i9} & \eta_{i10} & \eta_{i12} \\ \eta_{i9}& \eta_{i5} & \eta_{i11} & \eta_{i13} \\ \eta_{i10}  &  \eta_{i11} & \eta_{i7} & \eta_{i14} \\\eta_{i12} & \eta_{i13} & \eta_{i14} &  \eta_{i8} \end{pmatrix}$$

whose relation with the matrices $\log \boldsymbol D^2_i$ and $\boldsymbol T_i$ is simply $[\boldsymbol \Theta_i]_{jj}=[\log \boldsymbol D^2_i]_{jj}$, $j=1, \ldots, 4$, and  $[\boldsymbol \Theta_i]_{jk}= [\boldsymbol \Theta_i]_{kj}=  [\boldsymbol T_i]_{jk}$, $1 \leq k < j \leq 4$. 

For instance, consider the following model formula for the covariance matrix model:

```{r covmodelformula, echo=TRUE, eval=TRUE, include=TRUE}
my_k2 = 10
my_bs = "tp"
mformula <- c(mformula, list( 
                 Th_11 | Th_22 | Th_33 | Th_44 | Th_12 | Th_23 | Th_34  ~ dow + s(doy, k = my_k, bs = my_bs),
                 Th_11 ~ s(temp95_h17),
                 Th_22 ~ s(temp95_h18),
                 Th_33 ~ s(temp95_h19),
                 Th_44 ~ s(temp95_h20)))
```

As in the mean model, the covariance matrix model can be stated using a "Common" and a "Single-element" specification. However,  the l.h.s. of the model formula statement `y ~ x` targets the element of  $\boldsymbol \Theta$ to be modelled via the sintax   `Th_st ~ x`, where $s$ and $t$ are the indices targeting the elements of $\boldsymbol \Theta_i$.



Note that not specifying a model formula for a component of $\Theta$ that implies such element is automatically modelled via intercept. In contrast, the model formula for the mean vector components requires to be specified in any case.


The model is then fitted via 
```{r fit1, echo=TRUE, eval=TRUE, include=TRUE}
fit1 <- gam_scm(mformula, family = mvn_scm(d=4), data = GEF14_d4)
```

The resulting object `fit1` is of class `"scm"` and can be explored via

- `plot()`: to obtain the plot of the smooth effects on the linear predictor scale   
```{r gam plot, echo=TRUE, eval=FALSE, include = TRUE} 
plot(fit1, scale = FALSE, pages = 1) 
```

- `summary()`: summary of linear and smooth effects (`intercept = FALSE` avoids printing the intercepts)  
```{r summary, echo=TRUE, eval=FALSE, include = TRUE} 
summary(fit1, intercept = FALSE) 
```

- `residuals()`: raw residuals (`type = "response"`), uncorrelated standard normal residuals (default, `"deviance"`) and Pearson's residuals (`"pearson"`)  

```{r residuals, echo=TRUE, eval=FALSE}
head(residuals(fit1, type = "deviance"))
```

The `predict()` function allows to obtain:

- fitted values in terms of $\boldsymbol \eta$

```{r pred1, echo=TRUE, eval=FALSE} 
head(predict(fit1))
```

- fitted values in terms of mean vector, variances and correlations (Note that the first four columns corresponds to the mean vector, the columns 5-8 the variances and the last six columns correspon to the correlations following the row-wise ordering, that is (2,1), (3,1), (3,2), (4,1), (4,2) and (4,3)) 

```{r pred2, echo=TRUE, eval=FALSE} 
head(predict(fit1, type = "response")) 
``` 

- fitted values in terms of (mean vector, variances, covariances)
```{r pred3, echo=TRUE, eval=TRUE} 
fit1$family$put_cflag(FALSE) 
Sigma_pred <- predict(fit1, type = "response")
head(Sigma_pred)
``` 

The function `Sigma_mat()` allows to organise the covariance matrices in a list, for instance  
```{r pred4, echo=TRUE, eval=TRUE} 
Sigma_mat(Sigma_pred[,-c(1:d)])[[1]]
``` 

The **mgcViz** package by Fasiolo et al. (2020) allows to obtain the residual diagnostics and the accumulated local effects (ALEs) by Apley and Zhu (2020). 
For instance, the centered ALE of the variable `doy' on the variance of the net-demand at 5 p.m. is obtained via  
```{r , echo=TRUE, eval=TRUE, message = F} 
library(mgcViz)
``` 

```{r ALEs, echo=TRUE, eval=TRUE} 
Sigma_el <- SCM:::sel_elem(d)
var_y1 <- ALE(fit1, x = "doy", type = "response",
                      oind = Sigma_el(1, 1), center = 1)
plot(var_y1)
``` 





References
==========
- Apley, D. W. and Zhu, J. (2020) Visualizing the effects of predictor variables in black box supervised learning models. Journal of the Royal Statistical Society: Series B (Statistical Methodology), 82 (4), 1059 -- 1086.  

- Chiu, T. Y. M., Leonard, T. and Tsui, K.-W. (1996) The matrix-logarithmic covariance model. Journal of the American Statistical Association, 91 (433), 198 -- 210.    

-   Fasiolo, M., Nedellec, R., Goude, Y. and Wood, S.N., 2020. Scalable visualization methods for modern generalized additive models. Journal of computational and Graphical Statistics, 29 (1), 78--86.

- Hong, T., Pinson, P., Fan, S., Zareipour, H., Troccoli, A., and Hyndman, R.J. (2016). Probabilistic energy forecasting: global energy forecasting competition 2014 and beyond. International Journal of Forecasting, 32, 896 -- 913.

- Pourahmadi, M. (1999) Joint mean-covariance models with applications to longitudinal data: unconstrained parameterisation. Biometrika,  86 (3), 677 -- 690.

- Wood, S. N., Pya, N. and S\"{a}fken, B. (2016) Smoothing parameter and model selection for general smooth models (with discussion).  Journal of the American Statistical Association, 111 (516), 1548 -- 1575.

-  Wood, S. N. (2017) Generalized Additive Models: An Introduction with R, 2nd ed. Boca Raton, FL, USA: Chapman and Hall/CRC.

-  Wood, S. N. and Fasiolo, M.  (2017). A generalized Fellner-Schall method for smoothing parameter optimization with application to Tweedie location, scale and shape models.  Biometrics, 73 (4), 1071 -- 1081.
